\begin{acknowledgements}
My special appreciation and thanks go to my advisors Dr. Heinz Pernegger and Dr. Erich Griesmayer. Your support of my work has been invaluable. By accepting me to the Marie Curie TALENT programme you allowed me to embark on a life-changing and eye-opening journey, which will definitely have a major impact on me in the years to come. Your supervision in the past four years allowed me to grow professionally as well as personally.

I will take this opportunity to acknowledge that this work position at CERN was included in the Work Package 2 of the Marie Curie Early Initial Training Network Fellowship of the European Community’s Seventh Framework Programme under contract number (PITN-GA-2011-289161- TALENT). Furthermore, the work has been generously supported by CIVIDEC Instrumentation Gmbh. 

I would also like to thank Andrej Gori\v{s}ek, Marko Miku\v{z} and Marko Zavrtanik. Without your selfless support during my early days as a researcher I would not be where I am today. You gave me an insight into joys of particle physics research (i.e. playing around with expensive hardware), which gave rise to my increased interest in the field and my eventual involvement with CERN.

Next I would like to extend my gratitude to my dear colleagues in the work groups: Jennifer, Sonia, Karola, Timon, Christian, Christian, Branislav, Simon, Doug, Malte, Hendrik, Garrin, Chav, Christina and Pavel. You stood by my side in my days of sorrow when the measurement setup decided to punish me with abstract errors, you helped me fight the battles against segfaults, corrected my sloppy writing, had insightful and most welcome ideas whenever asked for them and, most importantly, you made sure we never ran out of much needed coffee and banter in the office.

The same goes for my fellow TALENT colleagues, it was an enjoyable three-year-long adventure and I am glad to have had the privilege to share it with you. 

Jacek, Ian, Florentina and Torsten, thank you for your patience and to allow me to learn from your experience as I was testing electronics, assembling detectors and building measurement setups.

A big shout-out to all my Geneva friends. Our friendship has left a big mark on my life and I will never forget the fun, the music and all the great moments we spent together.

I am deeply thankful to my family for their love and support. Vida, Roman, you always believed in me and encouraged me on the path I chose and for this I will be forever indebted to you. Miha, Klemen, Ana, Meta, Zala, Julija, you are the best. Finally, I would like to thank Ur\v{s}ka. You have walked the path alongside me and supported me and made these years together ever so enjoyable.

Matev\v{z}	

\end{acknowledgements}
